Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-29 paper/main.tex
link to supplement; fill supplement; reorder some parts

\subsubsection{Imaging-based Comparison Algorithm}
\label{subsec:imaging-based-comparison}
The preceding algorithm is performant and effective but can not be applied in the case when a
division event splits apart the bacterium since this would involve the comparison of two sets of
vertices with only one.
This configuration is displayed in figure~\ref{fig:mask-difference-metric} (E) where two distinct
bacterial configurations originate from the same initial state but contain different amounts of
bacteria.
This mismatch can happen, e.g. due to non-matching division times.
Thus a cell which is present in the simulation might not yet have a corresponding cell in the
experimental data at this time point yet (or vice-versa).
To explain why the previous developed direct comparison algorithm can not be reused on confined
intervals which do not include division events and for further details, we refer the interested
reader to the supplementary
material~\ref{sec:supplement-imaging-based-comparison-algorithm-details}.

@@

%###################################################################################################
\subsection{Details for the Imaging-Based Comparison Algorithm}
\label{sec:supplement-imaging-based-comparison-algorithm-details}

@@

\begin{figure}
    \centering
    \begin{tikzonimage}[width=0.5\textwidth]
        {docs/source/_static/fitting-methods/displacement-distribution.pdf}
        \node at (0, 1)[anchor=north west, black]{\textbf{A}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.5\textwidth]
        {docs/source/_static/fitting-methods/displacement-fit-over-time-covariance.pdf}
        \node at (0, 1)[anchor=north west, black]{\textbf{B}};
    \end{tikzonimage}\\
    \begin{tikzonimage}[width=0.5\textwidth]
        {docs/source/_static/fitting-methods/displacement-distr-x.pdf}
        \node at (0, 1)[anchor=north west, black]{\textbf{C}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.5\textwidth]
        {docs/source/_static/fitting-methods/displacement-distr-y.pdf}
        \node at (0, 1)[anchor=north west, black]{\textbf{D}};
    \end{tikzonimage}
    \caption{TODO}
\end{figure}

>>>>

\subsubsection{Imaging-based Comparison Algorithm}
\label{subsec:imaging-based-comparison}
The preceding algorithm is performant and effective but can not be applied in the case when a
division event splits apart the bacterium since this would involve the comparison of two sets of
vertices with only one.
This configuration is displayed in figure~\ref{fig:mask-difference-metric} (E) where two distinct
bacterial configurations originate from the same initial state but contain different amounts of
bacteria.
This mismatch can happen, e.g. due to non-matching division times.
Thus a cell which is present in the simulation might not yet have a corresponding cell in the
experimental data at this time point yet (or vice-versa).
To explain why the previous developed direct comparison algorithm can not be reused on confined
intervals which do not include division events and for further details, we refer the interested
reader to the supplementary
material~\ref{sec:supplement-comparison-algorithms}.

@@

\subsection{Further Details of the Extraction Algorithm}
\label{sec:supplement-extraction-algorithm}

The main text stated that the distribution of displacements which has been split up into parallel
and orthogonal components along the rod can be represented by a multivariate normal distribution.
We first describe in more detail how these distributions are obtained.
Consider the case where we have a set of exactly known vertices $\nu_{j,k}$ and a set of extracted
vertices $\nu_{j,k}'$.
We now calculate the displacements simply via
\begin{equation}
    \Delta \nu_{j,k} = \nu_{j,k} - \nu_{j,k}'.
\end{equation}
In order to calculate their positions with respect to the direction of the rod, we assign a
direction $d_{j,k}$ at each vertex of the known exact values.
At the tips of the rod this is done by assigning the direction of the closest edge while vertices in
the middle are calculated using a central difference approach.
\begin{align}
    d_{0,k} &= \frac{(\nu_{1,k}-\nu_{0,k})}{||\nu_1 - \nu_0||}\\
    d_{N_v,k} &= \frac{(\nu_{N_v,k}-\nu_{N_v-1,k})}{||\nu_{N_v} - \nu_{N_v-1}||}\\
    d_{j,k} &= \frac{(\nu_{j+1,k}-\nu_{j-1,k})}{||\nu_{j+1} - \nu_{j-1}||}
\end{align}
We can use these directions at each vertex to assign rotation matrices
\begin{equation}
    M_{j,l}^k = \begin{pmatrix}
        d_{j,0} & d_{j,1}\\
        -d_{j,1} & d_{j,0}
    \end{pmatrix}
\end{equation}
at each vertex and apply them to the displacements via
\begin{equation}
    (\Delta\nu_{j,k})_\text{directed} = \sum\limits_l M_{j,l}^k \Delta\nu_{j,l}.
\end{equation}
The resulting points are displayed in Figure~\ref{fig:supplement-extraction-algorithm} (A) which is
identical to the one in the main text.

The main text claims that this distribution is well described by a multimodal normal distribution
with diagonal covariance matrix.
Figure~\ref{fig:supplement-extraction-algorithm} (C,D) show slices of the full distribution along
the parallel and orthogonal axis.
Darker colors indicate earlier stages of the simulation while lighter colors indicate later stages.
Visually, we can already see that the distribution is quite narrow initially but widens with
increased simulation time.
To quantify this behaviour, we fit a multimodal normal distribution at each time point to the
distribution of displacements.
The main text has already discussed results for the variances $\sigma_{00}$ and $\sigma_{11}$.
Figure~\ref{fig:supplement-extraction-algorithm} (B) shows the trajectory of the covariance
$\sigma_{01}=\sigma_{10}$ which has been normalized to the value of $\sigma_{00}$.
Its small relative value justifies the statement in the main text and shows no correlation between
parallel and orthogonal displacement values.

\begin{itemize}
    \item for low resolution pictures this can be a problem
    \item in our benchmarking it is not important since we chose large resolution where such
        artifacts can not occur
\end{itemize}

\begin{figure}[H]
    \centering
    \begin{tikzonimage}[width=0.5\textwidth]
        {docs/source/_static/fitting-methods/displacement-distribution.pdf}
        \node at (0, 1)[anchor=north west, black]{\textbf{A}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.5\textwidth]
        {docs/source/_static/fitting-methods/displacement-fit-over-time-covariance.pdf}
        \node at (0, 1)[anchor=north west, black]{\textbf{B}};
    \end{tikzonimage}\\
    \begin{tikzonimage}[width=0.5\textwidth]
        {docs/source/_static/fitting-methods/displacement-distr-x.pdf}
        \node at (0, 1)[anchor=north west, black]{\textbf{C}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.5\textwidth]
        {docs/source/_static/fitting-methods/displacement-distr-y.pdf}
        \node at (0, 1)[anchor=north west, black]{\textbf{D}};
    \end{tikzonimage}
    \caption{TODO}
    \label{fig:supplement-extraction-algorithm}
\end{figure}

%###################################################################################################
\subsection{Details for the Imaging-Based Comparison Algorithm}
\label{sec:supplement-comparison-algorithms}

@@
